MS Technical Paper: \\ Placement Algorithms for Heterogeneous FPGAs

### 2025-04-16 latex/paper/paper.tex
move cell pattern explanation into architecture section

    The paper first begins by elaborating on general FPGA architecture and then specifically the Xilinx 7-Series architecture. 
    Then, the paper will elaborate on the FPGA design flow, then the role that the RapidWright API plays in the design flow. 
    We explain in detail each of these concepts for a broader audience as they provide much needed context for FPGA placement algorithms as a concept. 
    However, readers who are already familiar with these concepts can skip directly to the RapidWright API section.

@@

    Before any work can begin on an FPGA placer, it is necessary to understand both the objects being placed and the medium in which they are placed.
    Configurable logic devices have undergone significant evolution over the past four decades. 
    We will briefly review the evolution of configurable logic architecture starting in the 1970s and quickly work our way up to modern day FPGA architecture. 
    The journey began with the Programmable Logic Array (PLA) in the early 1970s. 
    The PLA implemented output logic using a programmable-OR and programmable-AND plane that formed a sum-of-products equation for each output through programmable fuses. 
    Around the same time, the Programmable Array Logic (PAL) was introduced. 
    The PAL simplified the PLA by fixing the OR gates, resulting in a fixed-OR, programmable-AND design, which sacrificed some logic flexibility to simplify its manufacture. 
    Figure \ref{fig:pla} shows one such PLA architecture.

@@

    Later in the same decade came the Complex Programmable Logic Device (CPLD), which took the form of an array of Configurable Logic Blocks (CLBs). 
    These CLBs were typically modified PAL blocks that included the PAL itself along with macrocells such as flip-flops, multiplexers, and tri-state buffers. 
    The CPLD functioned as an array of PALs connected by a central programmable switch matrix and could be programmed using a hardware description language (HDL) like VHDL. 
    Figure \ref{fig:cpld} shows one such CPLD architecture.

@@

    The mid-1980s saw the introduction of homogeneous FPGAs, which were built as a grid of CLBs. 
    Rather than using a central programmable switch matrix as in CPLDs, FPGAs adopted an island style architecture in which each CLB is surrounded on all sides by programmable routing resources, as shown in Figure \ref{fig:fpga}. 
    The first commercially viable FPGA, produced by Xilinx in 1984, featured 16 CLBs arranged in a 4x4 grid. 
    As FPGA technology advanced, CLBs were redesigned to use lookup tables (LUTs) instead of PAL arrays for greater logic density. 
    The capacity of an FPGA was often measured by how many logical elements or CLBs it offered, which grew from hundreds to thousands and now to hundreds of thousands of CLBs.

    This brings us to modern day FPGA architectures. 
    To meet the needs of increasingly complex designs, FPGA vendors introduced heterogeneous FPGAs. 
    In these devices, hard macros such as Block RAM (BRAM) and Digital Signal Processing (DSP) slices are integrated into the programmable logic fabric along with CLBs, like shown in Figure \ref{fig:fpga}. 
    This design enables the direct instantiation of common subsystems like memories and multipliers, without having to recreate them from scratch using CLBs. 
    Major vendors such as Xilinx and Altera now employ heterogeneous island-style architectures in their devices. 
    As designs become increasingly large and complex, FPGAs meet the demand by becoming increasingly heterogenous, incorporating a wider variety of hard macros into the fabric.

    \section{Xilinx 7-Series Architecture}

    \begin{figure*}[!ht]
        \centering
        \includegraphics[width=15.0cm]{figures/hierarchy_3.png}
        \caption{Architecture Hierarchy of a Xilinx FPGA}
        \label{fig:hierarchy_4}
    \end{figure*}

@@

    \textbf{SLICE Types.}\quad 
    Xilinx offers two variants of SLICE Sites: \textbf{SLICEL} and \textbf{SLICEM}.  
    \begin{itemize}
    \item \emph{SLICEL (SLICE-Logic):} Each LUT in a SLICEL can implement a 6-input Boolean function (with up to two outputs under certain input-sharing conditions). During logic synthesis, larger Boolean expressions exceeding six inputs are automatically split across multiple LUTs.
    \item \emph{SLICEM (SLICE-Memory):} SLICEM LUTs include all the features of a SLICEL but can also be configured as small distributed RAM (known as \emph{Distributed RAM}) or as shift registers. This local memory capability supplements the larger Block RAM (BRAM) cells and is beneficial for designs requiring numerous small RAM blocks.
    \end{itemize}

    \begin{figure}[!ht]
        \centering
        \includegraphics[width=7.5cm]{figures/slicel.png}
        \caption{A SLICEL Site in the Vivado device viewer}
        \label{fig:slicel}
    \end{figure}

    \begin{figure}[!ht]
        \centering
        \includegraphics[width=7.5cm]{figures/slicem.png}
        \caption{A SLICEM Site in the Vivado device viewer}
        \label{fig:slicem}
    \end{figure}

@@

    \textbf{CARRY4 for Arithmetic.}\quad
    Very often an HDL design will implement many adders, counters, subtractors, or comparators, all of which are based on binary addition. 
    They are so ubiquitous that every that in the 7-Series architecture, every SLICE features a \emph{CARRY4} BEL -- a 4-bit carry-lookahead (CLA) adder. 
    These CARRY4 blocks can be chained vertically across SLICEs for implementing wide adders efficiently.

    \textbf{Summary and Relevance to Placement.}\quad
    To summarize, Xilinx 7-Series FPGAs are structured into multiple Clock Regions, each divided into Tiles, which in turn contain Sites specialized for logic (CLB), DSP, RAM, or I/O. 
    Within each CLB Tile, the two SLICE Sites host the BELs (LUTs, FFs, and a CARRY4 adder). 
    Any placement tool must carefully ahdere to these hierarchy boundaries to ensure legal mappings that minimize wirelength and meet timing requirements.

@@

@@

        \item \textbf{Placement and Routing (Implementation):} 
            In \emph{placement}, each logical cell from the synthesized netlist is assigned to a physical location on the FPGA fabric. 
            For instance, LUTs and FFs go into specific \emph{BELs} within the device’s CLB sites, and specialized cells such as DSPs and Block RAMs must be placed in their corresponding tile types. 
            Next, \emph{routing} determines how signals are physically wired through the FPGA’s configurable interconnect network. 
            Modern tools often interleave these steps (e.g., placement-aware routing or routing-aware placement) to better meet timing and area objectives.

@@

    \begin{figure}[]
    \centering
    \includegraphics[width=7.5cm]{figures/design_flow.png}
    \caption{Typical FPGA design and verification workflow.}
    \label{fig:design_flow.png}
    \end{figure}

@@

    \textbf{RapidWright} is an open-source Java framework from AMD/Xilinx that provides direct access to the netlist and device databases used by vendor tools. 
    This framework positions itself as an additional workflow column, allowing users to intercept or replace parts of the standard design flow with custom optimization stages (see Figure~\ref{fig:vivado_dcps}).

@@

    \item \textbf{Interfacing with the Netlist and Device:} 
        An engineer can query the netlist to find specific resources (LUTs, FFs, DSPs, etc.) and then map or move them onto device sites. 
        This level of control is invaluable for research in custom placement, advanced packing techniques, or experimental routing algorithms.
    \end{itemize}

@@

\section{Simulated Annealing}
    With a basic understanding of FPGA architecture, design placement, and RapidWright, we have all the necessary pieces to implement our SA placer. 
    Here we outline in detail each substage of our implementation: PrePacking, Packing, and Placement.

@@

        In the context of implementing a custom \emph{Simulated Annealing} (SA) placer, \textbf{prepacking} is the first step in consolidating functionally related cells from the raw post-synthesis netlist into logical clusters that correspond more naturally to the physical FPGA structures. 
        We traverse the netlist (e.g., via RapidWright’s \texttt{edif} and \texttt{design} packages) to identify:

        \begin{itemize}
        \item \textbf{LUT-FF pairs:} 
            Most synchronous designs implement finite-state machines or pipelined datapaths, mapping their combinational logic into LUTs directly feeding FFs. 
            Clustering each LUT-FF pair under a single “prepacked object” facilitates usage of dedicated intra-SLICE connections, thereby reducing routing overhead.

        \item \textbf{CARRY4 chains:} 
            Arithmetic operations (adders, counters, comparators) often chain multiple \emph{CARRY4} cells vertically. 
            During prepacking, it is crucial to detect these chains so they remain grouped and placed in physically consecutive and vertical SLICEs for placement legality and optimal performance.

        \item \textbf{DSP cascades:} 
            In designs with large multiply-accumulate structures, DSP48E1 blocks can be used and cascaded along dedicated paths. 
            Identifying such cascades early helps the placer keep them localized, minimizing high-speed interconnect usage.
            Like carry chains, each DSP in a DSP cascades must be placed vertically, in physically consecutive and vertical DSP48E1 Sites.

        \item \textbf{Other specialized patterns:} 
            Various design-specific macros (e.g., BRAM-based FIFOs, shift-register logic) also benefit from being clustered, though these may be addressed in subsequent or more advanced stages.
        \end{itemize}

        These prepacked “super-cells” or “block groups” more closely match the physical site resources. 
        This strategy not only reduces the complexity of the placement search space but also helps ensure legality (e.g., shared control signals, shared carry nets). 
        The subsequent \emph{packing} and \emph{placement} steps can then operate on these higher-level groupings with improved efficiency.

>>>>

    The paper first begins by elaborating on general FPGA architecture and then specifically the Xilinx 7-Series architecture. 
    Then, the paper will elaborate on the FPGA design flow, then the role that the RapidWright API plays in the design flow. 
    We explain in detail each of these concepts for a broader audience as they provide much needed context for FPGA placement algorithms as a concept. 
    However, readers who are already familiar with these concepts can skip directly to the RapidWright API section \ref{sec:rapidwright_api} or to the Simulated Annealing section \ref{sec:simulated_annealing}.

@@

    Before any work can begin on an FPGA placer, it is necessary to understand both the objects being placed and the medium in which they are placed.
    Configurable logic devices have undergone significant evolution over the past four decades. 
    We will briefly review the evolution of configurable logic architecture starting in the 1970s and quickly work our way up to modern day FPGA architecture.

    \textbf{PLA:} \quad
    The journey began with the Programmable Logic Array (PLA) in the early 1970s. 
    The PLA implemented output logic using a programmable-OR and programmable-AND plane that formed a sum-of-products equation for each output through programmable fuses. 
    Around the same time, the Programmable Array Logic (PAL) was introduced. 
    The PAL simplified the PLA by fixing the OR gates, resulting in a fixed-OR, programmable-AND design, which sacrificed some logic flexibility to simplify its manufacture. 
    Figure \ref{fig:pla} shows one such PLA architecture.

@@

    \textbf{CPLD:} \quad 
    Later in the same decade came the Complex Programmable Logic Device (CPLD), which took the form of an array of Configurable Logic Blocks (CLBs). 
    These CLBs were typically modified PAL blocks that included the PAL itself along with macrocells such as flip-flops, multiplexers, and tri-state buffers. 
    The CPLD functioned as an array of PALs connected by a central programmable switch matrix and could be programmed using a hardware description language (HDL) like VHDL. 
    Figure \ref{fig:cpld} shows one such CPLD architecture.

@@

    \textbf{Homogeneous FPGA:} \quad 
    The mid-1980s saw the introduction of homogeneous FPGAs, which were built as a grid of CLBs. 
    Rather than using a central programmable switch matrix as in CPLDs, FPGAs adopted an island style architecture in which each CLB is surrounded on all sides by programmable routing resources, as shown in Figure \ref{fig:fpga}. 
    The first commercially viable FPGA, produced by Xilinx in 1984, featured 16 CLBs arranged in a 4x4 grid. 
    As FPGA technology advanced, CLBs were redesigned to use lookup tables (LUTs) instead of PAL arrays for greater logic density. 
    The capacity of an FPGA was often measured by how many logical elements or CLBs it offered, which grew from hundreds to thousands and now to hundreds of thousands of CLBs.

    \textbf{Heterogenous FPGA:} \quad 
    This brings us to modern day FPGA architectures. 
    To meet the needs of increasingly complex designs, FPGA vendors introduced heterogeneous FPGAs. 
    In these devices, hard macros such as Block RAM (BRAM) and Digital Signal Processing (DSP) slices are integrated into the programmable logic fabric along with CLBs, like shown in Figure \ref{fig:fpga}. 
    This design enables the direct instantiation of common subsystems like memories and multipliers, without having to recreate them from scratch using CLBs. 
    Major vendors such as Xilinx and Altera now employ heterogeneous island-style architectures in their devices. 
    As designs become increasingly large and complex, FPGAs meet the demand by becoming increasingly heterogenous, incorporating a wider variety of hard macros into the fabric.

    \section{Xilinx 7-Series Architecture}
    \label{sec:7_series}

    \begin{figure}[]
        \centering
        \includegraphics[width=5.0cm]{figures/hierarchy_5.png}
        \caption{Overview of Architecture Hierarchy of a Xilinx FPGA}
        \label{fig:hierarchy_5}
    \end{figure}

    \begin{figure}[]
        \centering
        \includegraphics[width=9.0cm]{figures/hierarchy_4.png}
        \caption{Detailed Architecture Hierarchy of a Xilinx FPGA}
        \label{fig:hierarchy_4}
    \end{figure}

@@

    \textbf{SLICE Types.}\quad 
    Xilinx offers two variants of SLICE Sites: \textbf{SLICEL} and \textbf{SLICEM}.  
    \begin{itemize}
        \item \emph{SLICEL (SLICE-Logic):} Each LUT in a SLICEL can implement a 6-input Boolean function (with up to two outputs under certain input-sharing conditions). During logic synthesis, larger Boolean expressions exceeding six inputs are automatically split across multiple LUTs.
        \item \emph{SLICEM (SLICE-Memory):} The SLICEM include all the features of a SLICEL but its LUTs can also be configured as small distributed RAM (known as \emph{Distributed RAM}) or as shift registers. This local memory capability supplements the larger Block RAM (BRAM) cells and is beneficial for designs requiring numerous small RAM blocks.
    \end{itemize}

    \begin{figure}[h]
        \centering
        \includegraphics[width=7.5cm]{figures/slicel.png}
        \caption{A SLICEL Site in the Vivado device viewer}
        \label{fig:slicel}
    \end{figure}

    \begin{figure}[h]
        \centering
        \includegraphics[width=7.5cm]{figures/slicem.png}
        \caption{A SLICEM Site in the Vivado device viewer}
        \label{fig:slicem}
    \end{figure}

@@

    ( Explain cell patterns here )

    \textbf{LUTs} \quad

    \textbf{FFs} \quad

    \textbf{LUT-FF Pairs} \quad

    \textbf{CARRY} \quad
        Very often an HDL design will implement many adders, counters, subtractors, or comparators, all of which are based on binary addition. 
        They are so ubiquitous that every that in the 7-Series architecture, every SLICE features a \emph{CARRY4} BEL -- a 4-bit carry-lookahead (CLA) adder. 
        These CARRY4 blocks can be chained vertically across SLICEs for implementing wide adders efficiently.

    \textbf{CARRY Chains} \quad

    \textbf{DSPs} \quad

    \textbf{DSP Cascades} \quad

    \textbf{BRAMs} \quad

    \textbf{Loose Single Cells} \quad

@@

    \begin{figure}[]
    \centering
    \includegraphics[width=7.5cm]{figures/design_flow.png}
    \caption{Typical FPGA design and verification workflow.}
    \label{fig:design_flow}
    \end{figure}

@@

        \item \textbf{Placement and Routing (Implementation):} 
            In \emph{placement}, each logical cell from the synthesized netlist is assigned to a physical location on the FPGA fabric. 
            For instance, LUTs and FFs go into specific \emph{BELs} within the device’s CLB sites, and specialized cells such as DSPs and Block RAMs must be placed in their corresponding tile types. 
            Next, \emph{routing} determines how signals are physically wired through the FPGA’s configurable interconnect network. 
            Modern tools often interleave these steps (e.g., fluid-placement routing or routing-aware placement) to better meet timing and area objectives.

@@

@@

    \textbf{RapidWright} is an open-source Java framework from AMD/Xilinx that provides direct access to the netlist and device databases used by vendor tools. 
    This framework positions itself as an additional workflow column, allowing users to intercept or replace stages of the standard design flow with custom optimization stages (see Figure~\ref{fig:vivado_dcps}).

@@

    \item \textbf{Interfacing with the Netlist and Device:} 
        An engineer can query the netlist to find specific resources (LUTs, FFs, DSPs, etc.) and then map or move them onto device sites. 
        This level of control over backend resources is necessary for research in custom placement, advanced packing techniques, or experimental routing algorithms.
    \end{itemize}

@@

\section{Simulated Annealing}
    \label{sec:simulated_annealing}
    With a basic understanding of FPGA architecture, design placement, and RapidWright, we have all the necessary pieces to implement our SA placer. 
    Here we outline in detail each substage of our implementation: PrePacking, Packing, and Placement. 
    Shown in Figure

    \begin{figure}[]
        \centering
        \includegraphics[width=8.0cm]{figures/edif_design_device.png}
        \caption{}
    \end{figure}

@@

        In the context of implementing a custom \emph{Simulated Annealing} (SA) placer, \textbf{prepacking} is the first step in consolidating functionally related cells from the raw post-synthesis netlist into logical clusters that correspond more naturally to the physical FPGA structures. 
        We traverse the raw post-synthesis netlist via RapidWright’s \texttt{edif} packages to identify LUT-FF Pairs, CARRY4 Chains, and DSP Cascades.

        \textbf{LUT-FF pairs:} \quad
            Most synchronous designs implement finite-state machines or pipelined datapaths, mapping their combinational logic into LUTs directly feeding FFs. 
            Clustering each LUT-FF pair under a single “prepacked object” facilitates usage of dedicated intra-SLICE connections, thereby reducing routing overhead.

        \textbf{CARRY4 chains:} \quad
            Arithmetic operations (adders, counters, comparators) often chain multiple \emph{CARRY4} cells vertically. 
            During prepacking, it is crucial to detect these chains so they remain grouped and placed in physically consecutive and vertical SLICEs for placement legality and optimal performance.

        \textbf{DSP cascades:} \quad
            In designs with large multiply-accumulate structures, DSP48E1 blocks can be used and cascaded along dedicated paths. 
            Identifying such cascades early helps the placer keep them localized, minimizing high-speed interconnect usage.
            Like carry chains, each DSP in a DSP cascades must be placed vertically, in physically consecutive and vertical DSP48E1 Sites.

        \textbf{Loose or Single Cells} \quad
            The remaining cells 
            Various design-specific macros (e.g., BRAM-based FIFOs, shift-register logic) also benefit from being clustered, though these may be addressed in subsequent or more advanced stages.

        This strategy not only reduces the complexity of the placement search space but also helps ensure legality (e.g., shared control signals, shared carry nets). 
        The subsequent \emph{packing} and \emph{placement} steps can then operate on these higher-level groupings with improved efficiency.

    \subsection{Packing}
        \label{subsec:packing}

    \subsection{Placement}
        \label{subsec:placement}
        Up until now we have only organized the logical \texttt{Cells} into \texttt{SiteInsts}. 
        This is where simulated annealing actually begins.
